\propertitle

### 2019-12-07 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-07T14:46:25+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

Here is a summary derivation of the maximum-entropy distribution for the
larger population constrained by a set of factorial moments. For further details
see Porta Mana \etal\ \citey{portamanaetal2015}.

@@

First of all the sampling relation. We have a set of $N$ units, $\yA$ of
which have some specific property, and we sample in an unknown way $n$ of
the $N$ units. The probability that $\ya$ of the $n$ sampled units have that
property is then given by the hypergeometric distribution
\begin{equation}
  \label{eq:conditional_hypergeometric}
  \yG_{\ya\yA} \defd \pf(\ya \|\yA, n,N)=
  \binom{\yA}{\ya}\binom{N-\yA}{n-\ya}\binom{N}{n}^{-1}
\equiv  \binom{n}{\ya}\binom{N-n}{\yA-\ya}\binom{N}{\yA}^{-1}
\end{equation}
typical of \enquote{drawing without replacement}
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}[\sects~2.1,
3.2]{jeffreys1939_r1983}.
In the following leave $n$, $N$ implicit in the conditional. If we are
uncertain about the number $\yA$, with belief $P(\yA)$, then by the theorem
of total probability our belief about $\ya$ is
\begin{equation}
  \label{eq:belief_sample_unknown_A}
  p(\ya)= \sum_{\yA} \yG_{\ya\yA} \;P(\yA).
\end{equation}
In our case the units are neurons, the set is the larger population, and the
property is their being active in a specific time bin.

@@

The result has the standard exponential-family form
\begin{equation}
  \label{eq:app_maxent_pop}
  \begin{gathered}
  \px(\yA)  = \frac{1}{Z(\yll)}\;
  g(\yA)\;
  \exp\Biggl[\sum_{m}\yl_{m}\;
  \binom{\yA}{m}\binom{N}{m}^{-1}
  \Biggr],
  \\
  Z(\yll)\defd \sum_{\yA}g(\yA)\;
  \exp\Biggl[\sum_{m}\yl_{m}\;
  \binom{\yA}{m}\binom{N}{m}^{-1}
  \Biggr],
\end{gathered}
\end{equation}
where $g(\yA)$ is the reference distribution and $\yll\defd (\yl_{m})$ are
the Lagrange multipliers, satisfying the implicit
constraint equations~\eqref{eq:final_constraint}:
\begin{equation}
  \label{eq:constraint_eq_lambda}
  \sum_{\yA}
  \tbinom{\yA}{m}\tbinom{N}{m}^{-1}\;
  \tfrac{1}{Z(\yll)}\;
  g(\yA)\;
  \exp\biggl[\sum_{m}\yl_{m}\;
  \tbinom{\yA}{m}\tbinom{N}{m}^{-1}
  \biggr]
=
  \sum_{\ya}
  \tbinom{\ya}{m} \tbinom{n}{m}^{-1} \; \yff_{\ya},\quad
  m\in M.
\end{equation}

The reference distribution $g(\yA)$ represents our pre-data beliefs about
the activity levels $\yA$. We know that the majority of neurons in a brain
area are rarely simultaneously active within a window of some milliseconds,
so we could choose a distribution with slightly higher weights on low
values of $\yA$. On the other hand, considering the number of ways in which
$\yA$ out of $N$ neurons can be simultaneously active would suggest the
multiplicity distribution proportional to $\binom{N}{\yA}$. It turns out
that our results of \sects~\ref{sec:application}--\ref{sec:marginalization}
are actually quite insensitive to the choice between these two possible
reference distributions, or even a uniform reference distribution.

@@

Let's ask how much more probable is the
sufficiency of one set with respect to the other, conditional on our data
$\yf$:
\begin{equation}
  \label{eq:ratio_prob_suffic}
  \pf(M'' \| \yf)/
  \pf(M' \| \yf).
\end{equation}
Now, the probability of observing activity $\ya$ in the sample at any time
bin is the sample marginal of the maximum-entropy distribution for the larger
population, owing to the excheangeability assumption implicit in the
maximum-entropy method:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya_{t} \| M) = \sum_{\yA}\yG_{\ya_{t}\,\yA}\;\px(\yA \| M).
\end{equation}
The probability of observing one sequence $(\ya_{t})$ with frequencies
$\yf$ is therefore
\begin{equation}
  \label{eq:prob_frequencies}
  \prod_{t=1}^{T} \pxx(\ya_{t} \| M) \equiv
  \prod_{\ya=0}^{n} \pxx(\ya \| M)^{T\,\yff_{\ya}} \equiv
  \prod_{\ya=0}^{n}\biggl[\sum_{\yA}\yG_{\ya\yA}\;\px(\yA \| M)\biggr]^{T\,\yff_{\ya}}.
\end{equation}
The probability of observing the frequencies $\yf$ is obtained multiplying this
by their multiplicity factor, the multinomial coefficient
\begin{equation}
  \label{eq:multiplicity_factor}
  \binom{T}{T\yf} \defd \frac{T!}{\prod_{\ya}(T\,\yff_{\ya})!}
  \approx \prod_{\ya}{\yff_{\ya}}^{-T\,\yff_{\ya}},
\end{equation}
the last expression coming from Stirling's approximation
\citep[Lemma~2.2]{csiszaretal2004b}.
If we assign equal pre-data probabilities to the two hypotheses $M'$ and $M''$,
each probability in the ratio~\eqref{eq:ratio_prob_suffic} then becomes, by
Bayes's theorem,
\begin{multline}
  \label{eq:prob_proportional_relentropy}
  \pf(M \| \yf) \propto
  \pf(\yf \| M) \times\text{\small const} \propto
  \binom{T}{T\yf}\,  \prod_{\ya}\biggl[\sum_{\yA}\yG_{\ya\yA}\px(\yA \| M)\biggr]^{
    T\,\yff_{\ya}}
   \approx{}\\
  \prod_{\ya}{\yff_{\ya}}^{-T\,\yff_{\ya}}
  \times
  \prod_{\ya}\biggl[\sum_{\yA}\yG_{\ya\yA}\px(\yA \| M)\biggr]^{
    T\,\yff_{\ya}}.
\end{multline}
The logarithm of the probability above is easily seen to be the number of
bins $T$ multiplied by relative entropy between the frequency distribution
$\yf$ and the sample marginal of the maximum-entropy distribution.

Thus, the difference~\eqref{eq:measure_suffic} is the logarithm of the
probability ratio~\eqref{eq:ratio_prob_suffic}. The exponential of the
difference~\eqref{eq:measure_suffic} tells us how much more probable is the
set $M''$ to be sufficient than the set $M'$.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-07T15:36:17+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

Here is a summary derivation of the maximum-entropy distribution for the
larger population constrained by a set of factorial moments.%  For further details
% see Porta Mana \etal\ \citey{portamanaetal2015}.

@@

First of all the sampling relation. We have a population of $N$ units --
neurons -- $\yA$ of which have some specific property -- being active in a
specific time bin -- and we sample in an unknown way $n$ of the $N$ units.
The probability that $\ya$ of the $n$ sampled neurons are active is then
given by the hypergeometric distribution
\begin{equation}
  \label{eq:conditional_hypergeometric}
  \yG_{\ya\yA} \defd \pf(\ya \|\yA, n,N)=
  \binom{\yA}{\ya}\binom{N-\yA}{n-\ya}\binom{N}{n}^{-1}
\equiv  \binom{n}{\ya}\binom{N-n}{\yA-\ya}\binom{N}{\yA}^{-1}
\end{equation}
typical of \enquote{drawing without replacement}
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}[\sects~2.1,
3.2]{jeffreys1939_r1983}. In the following we leave $n$, $N$ implicit in
the conditional. If we are uncertain about the number $\yA$, with belief
$P(\yA)$, then by the theorem of total probability our belief about $\ya$
is
\begin{equation}
  \label{eq:belief_sample_unknown_A}
  p(\ya)= \sum_{\yA} \yG_{\ya\yA} \;P(\yA).
\end{equation}

@@

The result has the standard exponential-family form
\begin{equation}
  \label{eq:app_maxent_pop}
  \begin{gathered}
  \px(\yA)  = \frac{1}{Z(\yll)}\;
  r_{\yA}\;
  \exp\Biggl[\sum_{m}\yl_{m}\;
  \binom{\yA}{m}\binom{N}{m}^{-1}
  \Biggr],
  \\
  Z(\yll)\defd \sum_{\yA} r_{\yA}\;
  \exp\Biggl[\sum_{m}\yl_{m}\;
  \binom{\yA}{m}\binom{N}{m}^{-1}
  \Biggr],
\end{gathered}
\end{equation}
where $(r_{\yA})$ is the reference distribution and $\yll\defd (\yl_{m})$ are
the Lagrange multipliers, satisfying the implicit
constraint equations~\eqref{eq:final_constraint}:
\begin{equation}
  \label{eq:constraint_eq_lambda}
  \sum_{\yA}
  \tbinom{\yA}{m}\tbinom{N}{m}^{-1}\;
  \tfrac{1}{Z(\yll)}\;
  r_{\yA}\;
  \exp\biggl[\sum_{m}\yl_{m}\;
  \tbinom{\yA}{m}\tbinom{N}{m}^{-1}
  \biggr]
=
  \sum_{\ya}
  \tbinom{\ya}{m} \tbinom{n}{m}^{-1} \; \yff_{\ya},\quad
  m\in M.
\end{equation}

The reference distribution $(r_{\yA})$ represents our pre-data beliefs
about the activity levels $\yA$. We know that the majority of neurons in a
brain area are rarely simultaneously active within a window of some
milliseconds, so we could choose a distribution with slightly higher
weights on low values of $\yA$. On the other hand, considering the number
of ways in which $\yA$ out of $N$ neurons can be simultaneously active
would suggest the multiplicity distribution proportional to
$\binom{N}{\yA}$. It turns out that our results of
\sects~\ref{sec:application}--\ref{sec:marginalization} are insensitive to
the choice between these two possible reference distributions, or even a
uniform reference distribution.

For a derivation from a Bayesian perspective see
appendix~\ref{sec:inference_prob}.

@@

Let's ask how much more probable is the
sufficiency of one set with respect to the other, conditional on our data
$\yf$:
\begin{equation}
  \label{eq:ratio_prob_suffic}
  \frac{p(M'' \| \yf)}{ p(M' \| \yf)}.
\end{equation}
The probability of observing activity $\ya$ in the sample at any time bin
is the sample marginal of the maximum-entropy distribution for the larger
population, owing to the excheangeability assumption implicit in the
maximum-entropy method:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya_{t} \| M) = \sum_{\yA}\yG_{\ya_{t}\,\yA}\;\px(\yA \| M).
\end{equation}
The probability of observing one sequence $(\ya_{t})$ with frequencies
$\yf$ is therefore
\begin{equation}
  \label{eq:prob_frequencies}
  \prod_{t=1}^{T} \pxx(\ya_{t} \| M) \equiv
  \prod_{\ya=0}^{n} \pxx(\ya \| M)^{T\,\yff_{\ya}} \equiv
  \prod_{\ya=0}^{n}\biggl[\sum_{\yA}\yG_{\ya\yA}\;\px(\yA \| M)\biggr]^{T\,\yff_{\ya}}.
\end{equation}
The probability of observing the frequencies $\yf$ is obtained multiplying this
by their multiplicity factor, the multinomial coefficient
\begin{equation}
  \label{eq:multiplicity_factor}
  \binom{T}{T\yf} \defd \frac{T!}{\prod_{\ya}(T\,\yff_{\ya})!}
  \approx \prod_{\ya}{\yff_{\ya}}^{-T\,\yff_{\ya}},
\end{equation}
the last expression coming from Stirling's approximation
\citep[Lemma~2.2]{csiszaretal2004b}.
If we assign equal pre-data probabilities to the two hypotheses $M'$ and $M''$,
each probability in the ratio~\eqref{eq:ratio_prob_suffic} then becomes, by
Bayes's theorem,
\begin{multline}
  \label{eq:prob_proportional_relentropy}
  \pf(M \| \yf) \propto
  \pf(\yf \| M) \times\text{\small const} \propto
  \binom{T}{T\yf}\,  \prod_{\ya}\biggl[\sum_{\yA}\yG_{\ya\yA}\px(\yA \| M)\biggr]^{
    T\,\yff_{\ya}}
   \approx{}\\
  \prod_{\ya}{\yff_{\ya}}^{-T\,\yff_{\ya}}
  \times
  \prod_{\ya}\biggl[\sum_{\yA}\yG_{\ya\yA}\px(\yA \| M)\biggr]^{
    T\,\yff_{\ya}}.
\end{multline}
The logarithm of the probability above is easily seen to be the number of
bins $T$ multiplied by relative entropy between the frequency distribution
$\yf$ and the sample marginal of the maximum-entropy distribution.

Thus, the difference~\eqref{eq:measure_suffic} is the logarithm of the
probability ratio~\eqref{eq:ratio_prob_suffic}. The exponential of the
difference~\eqref{eq:measure_suffic} tells us how much more probable the
hypothesis that the set $M''$ is sufficient than the hypothesis that the
set $M'$ is.
